---
title: "PlottingCode"
output: html_document
---
### Load the `ggplot2` library (for later plotting) and to access the `diamonds` dataset
```{r}
library(ggplot2)
```

### Quick glance at the data
```{r}
str(diamonds) # diamonds was loaded automatically when we imported ggplot2
```

# Generate a scatterplot in `base R`
```{r}
plot(x = diamonds$carat, y = diamonds$price)
```

## Change the shape, color, and size of the points
```{r}
plot(
  x = diamonds$carat,
  y = diamonds$price,
  pch = 21, # shape
  col = 'chocolate', # color
  cex = 3 # size
)
```
<br>
## Set the range of the x and y axes, add axis and plot titles and a legend.
```{r}
plot(
  x = diamonds$carat,
  y = diamonds$price,
  pch = 18, # shape
  col = ifelse(diamonds$cut == 'Fair', 'chocolate', 'blue'),
  cex = 3, # size
  xlim = c(min(diamonds$carat), 1), # x-axis range
  ylim = c(min(diamonds$price), 10000), # y-axis range
  xlab = 'Carat',
  ylab = 'Price',
  main = 'This Is The Plot Title!' # set the plot title
)
legend(
  x = 0.2, # x-coordinate on the graph of the center of the legend
  y = 9000, # x-coordinate on the graph of the center of the legend
  legend = c('Fair', 'Better'), # the labels for the legend
  col = c('chocolate', 'blue'), # colors in the same order as the above
  pch = 18 # all same shape
)
```
<br>
## Adding a regression line
To add a regression line, we first fit a model using `lm`, and then supply that model
(after generating the scatterplot) to the function `abline`.
```{r}
mod <- lm(diamonds$price ~ diamonds$carat)
# Generate the plot again
plot(
  x = diamonds$carat,
  y = diamonds$price,
  pch = 18, # shape
  col = ifelse(diamonds$cut == 'Fair', 'chocolate', 'blue'),
  cex = 3, # size
  xlim = c(min(diamonds$carat), 1), # x-axis range
  ylim = c(min(diamonds$price), 10000), # y-axis range
  xlab = 'Carat',
  ylab = 'Price',
  main = 'This Is The Plot Title!' # set the plot title
)
legend(
  x = 0.2, # x-coordinate on the graph of the center of the legend
  y = 9000, # x-coordinate on the graph of the center of the legend
  legend = c('Fair', 'Better'), # the labels for the legend
  col = c('chocolate', 'blue'), # colors in the same order as the above
  pch = 18 # all same shape
)
# Add the model to the plot:
abline(reg = mod)
```
<br>
The regression line can be customized using parameters that govern the aesthetics
of lines. These include (but are not limited to):


* `lty`: the type of the line, e.g., solid, dotted, dashed \n
* `lwd`: width of line \n
<br>
`col` is the same as for modifying points.
```{r}
plot(
  x = diamonds$carat,
  y = diamonds$price,
  pch = 18, # shape
  col = ifelse(diamonds$cut == 'Fair', 'chocolate', 'blue'),
  cex = 3, # size
  xlim = c(min(diamonds$carat), 1), # x-axis range
  ylim = c(min(diamonds$price), 10000), # y-axis range
  xlab = 'Carat',
  ylab = 'Price',
  main = 'This Is The Plot Title!' # set the plot title
)
legend(
  x = 0.2, # x-coordinate on the graph of the center of the legend
  y = 9000, # x-coordinate on the graph of the center of the legend
  legend = c('Fair', 'Better'), # the labels for the legend
  col = c('chocolate', 'blue'), # colors in the same order as the above
  pch = 18 # all same shape
)
# Add the model to the plot:
abline(reg = mod, lwd = 4, lty = 3, col ='grey')
```

For a complete overview of all the graphical parameters (such as `col` and `lty`) and
what they control on your graph, run:
```{r}
?`graphical parameters`
```

# `ggplot2`
`ggplot2` or "ggplot" operates by building a graph a layer at a time, beginning with
the square plain white or grey image on which your data will be projected, up to the
actual data points, and even additional visualizations such as regression lines that
are computed directly from the data during plotting
<br>
`Pro-tip: layers in ggplot are seperated by the "+" operator` \n

### A quick demonstration of layers:
```{r}
ggplot()
```
<br>
We render just a plotting region, but the data isn't mapped onto it anywhere or in anyway
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price))
```

Here we have mapped variables onto the axes, and as such, they have adopted (automatically)
the scale/range of those variables that `ggplot` has computed for us internally, but we
still have not mapped the data to an *geometric* shape.

**Here we map the data on the x and y axes to be represented geometrically as a point:**
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point() # note, geom_point inherits the mapping that we input to ggplot() above
```

As you may have started to pick up, ggplot works by mapping data to different elements
of each layer that you add. The basic breakdown is as follows:


* Layers are added sequentially, separated by the "+" \n
* Mapping a variable to a given aesthetic element is acheived by setting the 
`mapping` argument equal to a call to `aes`, in which the variable is assigned \n
* `geom_` functions determine how the data will be represented geometrically,
and these shape different parameters need to be set inside `aes`

## Recreating the regression plot above in `ggplot2`
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = as.factor(cut == 'Fair')),
             shape = 18, size = 3) +
  xlab('Carat') +
  ylab('Price') +
  scale_color_manual(values = c('blue', 'chocolate'), labels = c('Better', 'Fair')) +
  guides(color = guide_legend(title = '', reverse = T)) +
  coord_cartesian(xlim = c(min(diamonds$carat), 1), ylim = c(0, 10000)) +
  theme(legend.position = c(.2, .7)) +
  geom_smooth(method = 'lm', 
              formula = 'y ~ x',
              se = F,
              color = 'grey', 
              size = 1.4, 
              linetype = 2) +
  # Some theme options to make it look more like a base R plot:
  theme(plot.title = element_text(hjust = .5), text = element_text(size = 12)) +
  theme(panel.border = element_rect(color = 'black', fill = NA, size = .9)) +
  theme(panel.background = element_rect(color = 'black', fill = 'white')) +
  theme(panel.grid = element_line(color = 'lightgrey', size = .2)) +
  theme(plot.background = element_rect(fill = 'white')) +
  theme(axis.text = element_text(size = 10, color = 'black')) +
  # We can add a title too
  ggtitle(label = 'This Is the Plot Title!')
```

A huge advantage of `ggplot`'s use of the mapping and layering technique is that
you can easily map continuous values to sets of points.

**Here we color code by price itself, a continuous variable **
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = price),
             shape = 18, size = 3) +
  xlab('Carat') +
  ylab('Price') +
  #scale_color_manual(values = c('blue', 'chocolate'), labels = c('Better', 'Fair')) +
  guides(color = guide_colorbar(title = '', reverse = T)) +
  #coord_cartesian(xlim = c(min(diamonds$carat), 1), ylim = c(0, 10000)) +
  theme(legend.position = c(.1, .7)) +
  geom_smooth(method = 'lm', 
              formula = 'y ~ x',
              se = F,
              color = 'grey', 
              size = 1.4, 
              linetype = 2) +
  # Some theme options to make it look more like a base R plot:
  theme(plot.title = element_text(hjust = .5), text = element_text(size = 12)) +
  theme(panel.border = element_rect(color = 'black', fill = NA, size = .9)) +
  theme(panel.background = element_rect(color = 'black', fill = 'white')) +
  theme(panel.grid = element_line(color = 'lightgrey', size = .2)) +
  theme(plot.background = element_rect(fill = 'white')) +
  theme(axis.text = element_text(size = 10, color = 'black')) +
  # We can add a title too
  ggtitle(label = 'This Is the Plot Title!')
```
<br>

**We can also easily view effects between groups by mapping a grouping variable**
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = price),
             shape = 18, size = 3) +
  xlab('Carat') +
  ylab('Price') +
  geom_smooth(method = 'lm', 
              formula = 'y ~ x',
              se = F,
              color = 'grey', 
              size = 1.4, 
              linetype = 2) +
  # Some theme options to make it look more like a base R plot:
  theme(plot.title = element_text(hjust = .5), text = element_text(size = 12)) +
  theme(panel.border = element_rect(color = 'black', fill = NA, size = .9)) +
  theme(panel.background = element_rect(color = 'black', fill = 'white')) +
  theme(panel.grid = element_line(color = 'lightgrey', size = .2)) +
  theme(plot.background = element_rect(fill = 'white')) +
  theme(axis.text = element_text(size = 10, color = 'black')) +
  # We can add a title too
  ggtitle(label = 'This Is the Plot Title!') +
  # Assign the grouping variable
  facet_wrap(~cut, scales = 'free') +
  # And we can map the points to a new color pallette
  scale_color_gradient(low = 'gold', high = 'chocolate')
```

